---
title: "PCA"
author: 
  - Costa, W. G.^[Weverton Gomes da Costa, Pós-Doutorando, Departamento de Estatística - UFV, wevertonufv@gmail.com]
date: "`r Sys.Date()`"
site: workflowr::wflow_site
url: https://wevertongomescosta.github.io/GitHub-with-RStudio-LICAE
output:
  workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Principal Components analysis with Iris data  

## Collecting data

```{r Collecting data}
data <- iris
```
<br>

## Preparing data for the principal components analysis (PCA)

let's prepare this prepare this data to plot some boxplot of all the four traits, for that you will need the function `melt` of the `reshape2` package and the tidyverse package.

```{r Melting data}
install.packages("reshape2", repos = "https://cloud.r-project.org")
library(reshape2); library(tidyverse)

dataMelted <- data %>% reshape2::melt(data = .,
                                      id.vars = "Species",
                                      variable.name = "trait",
                                      value.name = "y")
head(dataMelted)
```
<br>

Great, now we have the data at the format to make boxplot from all traits at the same code line. so lets keep moving. For that we will use `ggplot2` package.

```{r Iris traits Box plots}
dataMelted %>% ggplot(aes(x = Species, y = y, fill = Species)) +
     geom_boxplot() + facet_wrap(~trait, scales = "free_y") +
     theme(legend.position = "none")
```
<br>

Great data, we can see a lot of differences between the `Species` for these traits.
It seems that we may have some correlation between `Petal Length` and `Width`. We also have different amplitude for these traits this will certainly results in different `phenotypic variance` between the traits, so we need to `scale` these traits before the PCA.

```{r Scaling the data}
DataSc <- data %>% select(-Species) %>%
     scale(x = ., center = TRUE, scale = TRUE) %>%
     as.data.frame() %>% 
     mutate(Species = data$Species)
head(DataSc)
```
<br>

### Principal Component Analysis (PCA)

So let's proceed for the PCA analysis, here we will use the `prcomp` function from `R status` package, so no need to call any package.

```{r Iris PCA analysis}
PCA <- prcomp(DataSc %>% select(-Species))
```
<br>

### Saving results

Let's save the important results in objects, so we could make some graphs with them.
<br>

**1. Accumulate percent of the total phenotypic variance explained by the principal components (PC)**

```{r Accumulate variance PCs}
Perc <- 100 * PCA$sdev^2 / sum(PCA$sdev^2)

PercAc <- as.vector(rep(NA, times = length(Perc)))
for(i in 1:length(Perc)) {
  PercAc[i] <- sum(Perc[1:i])
  names(PercAc)[i] <- i
}
names(PercAc) <- c("PC1", "PC2", "PC3", "PC4")
PercAc
```

Oh these data are high correlated.

<br>

**2. Correlations of the traits with the principal components (PC)**
```{r Collect correlations PC and traits}
CorTraits <- PCA$rotation
rownames(CorTraits) <- c("SepLen", "SepWid", "PetLen", "PetWid")
CorTraits

LabelsPCA <- CorTraits %>% as.data.frame %>%
     mutate(PC1 = PC1 + 0.15, .keep = "unused")
```

<br>

**3. Individuals scores for the principal components (PC)**
```{r}
ScoresSpecies <- PCA$x %>%
     as.data.frame %>% 
     mutate(Species = data$Species)

head(ScoresSpecies)
```

Great we got what we need to create our figures.

<br>

### Figures

The first figure will be a barplot of the accumulated variances explained by the `PC`.
We will use the color red the PC selected to use at the next figures.

```{r Barplot PCs}
barplot(PercAc, main = "Variance explained by PCA",
        ylab = "Cumulative variance (%)", xlab = "Number of retained PCs",
        col = c("red", "red", "gray", "gray", "gray"))
```

<br>

R markdown allows us to hide the code that create the figure, this could be done adding the argument `echo = FALSE` inside the curly brackets at the chunk. Using `echo` argument will print just the result of you chunk, link below.  

```{r Barplot PCs2, echo = FALSE}
barplot(PercAc, main = "Variance explained by PCA",
        ylab = "Cumulative variance (%)", xlab = "Number of retained PCs",
        col = c("red", "red", "gray", "gray", "gray"))
```

<br>

The last figure will be a scatter plot of the individuals with their score for the first two PCs with the correlation of the traits with the first two PCs.

```{r Scatter plot PCA}
ggplot(data = ScoresSpecies, aes(x = PC1, y = PC2, color = Species)) +
  geom_point() + geom_rug(alpha = 0.2, size = 1.5) +
  geom_segment(mapping = aes(x = 0, xend = 3*PC1, y = 0, yend = 3*PC2),
               colour = "red",
               data = CorTraits %>% as.data.frame,
               arrow = arrow(type = "closed",
                             length = unit(0.2,units = "cm"))) +
  geom_text(mapping = aes(x = PC1*3, y = PC2*3, label = rownames(LabelsPCA)),
            data = LabelsPCA, colour = "black") + 
  theme_bw() +
  xlab(paste("PC1 - ", round(Perc[1], digits = 2), "%", sep = "")) +
     ylab(paste("PC2 - ", round(Perc[2], digits = 2), "%", sep = ""))
```
<br>

This is the final results of the PC. Mostly of the variance explained by the 1˚PC is due to the between species `Setosa` Vs `Versicolor` and `Virginica`. The 2˚PC just explain variance within the species. Also the traits `Petal Length`, `Petal Width` and `Sepal Length` could be used to discriminate the species.
